The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-04 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Natashas comments

\title[The MWA radio environment]{The low-frequency environment of the Murchison Widefield Array: radio frequency interference analysis and mitigation}

@@

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the future Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV). After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data. Some advise is suggested to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.

@@

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 80--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is built in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

Recently, experiments to use the moon as a calibrator for EoR experiments have shown that the reflection of the moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially isolated to the centre of the observed disk caused by the moon, and the high resolution of LOFAR allows separating the reflected and intrinsic power from the moon \citep{vedantham-2014-todo}. In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

In this paper, we will describe the mitigation strategy implemented for the MWA, and show examples of RFI found and results of the RFI detection. Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved. This strategy is applied to 63~h of MWA data. In Sect.~\ref{ch:data-description}, these data are described. Sect.~\ref{ch:detection-results} presents examples of RFI that were found in these data, as well as the effectivity of the RFI detection. In Sect.~\ref{ch:lofar-comparison}, the results are compared to a previously performed LOFAR RFI survey. Finally, in Sect.~\ref{ch:conclusions-and-discussion} conclusions are drawn and discussed.

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in data processing. It is important to perform initial RFI detection at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

\begin{figure}
\begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data. In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimise quantisation noise. The 1.28~MHz pattern is due to the imperfect poly-phase filter.
}
\label{fig:bandpass}
\end{center}
\end{figure}

To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customising its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions, and varying fields of view. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency directions. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ on one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging for getting accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimising reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not significantly cause any time or bandwidth decorrelation up to the first beam null. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $uvw$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~MHz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data are lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV. The horizontal gray line represents the false-positives rate of the RFI detection.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

\section{Detection results} \label{ch:detection-results}
The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citep{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

The residual noise levels after flagging can be used to validate whether the flagged data are free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, DTV interference has to be detected with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result of the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from digital TV (DTV), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DTV frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DTV bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. }
\label{fig:dvb}
\end{figure*}

Because significant DTV interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DTV transmitters are terrestrial, and the fact that only this night observes the DTV, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

>>>>

\title[The MWA radio environment]{The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation}

@@

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the future Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data, and another 0.14\% is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV). After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data. Some advise is given to ensure the SKA is robust to RFI, which includes: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

\section{Introduction}
Recent years have seen growth in the impact of radio-frequency interference (RFI) on radio astronomy, due to increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.

@@

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR exists in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially isolated to the centre of the observed disk caused by the Moon, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}. In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

In this paper, we will describe the mitigation strategy implemented for the MWA, and show examples of RFI found and results of the RFI detection. Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved. This strategy is applied to 63~h of MWA data. In Sect.~\ref{ch:data-description}, these data are described. Sect.~\ref{ch:detection-results} presents examples of RFI that were found in these data, as well as the efficacy of the RFI detection. In Sect.~\ref{ch:lofar-comparison}, the results are compared to a previously performed LOFAR RFI survey. Finally, in Sect.~\ref{ch:conclusions-and-discussion}, conclusions are drawn and discussed.

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in data processing. It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

\begin{figure}
\begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
\caption{Correlator output RMS with respect to frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data. In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimise quantisation noise. The 1.28-MHz pattern is due to the imperfect poly-phase filter.
}
\label{fig:bandpass}
\end{center}
\end{figure}

To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows the customisation of its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions, and varying fields of view. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency directions. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ on one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimising reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995} and/or \textsc{wsclean} \citep{offringa-wsclean-2014}, the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that pre-runs \textsc{cotter} on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not cause any significant time or bandwidth decorrelation up to the first beam null. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to remove completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~MHz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data are lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV. The horizontal gray line represents the false-positives rate of the RFI detection.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citep{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

The residual noise levels after flagging can be used to validate whether the flagged data are free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, DTV interference has to be detected with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result of the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from digital TV (DTV), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DTV frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DTV bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well-flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. }
\label{fig:dvb}
\end{figure*}

Because significant DTV interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DTV transmitters are terrestrial, and the fact that only this night observes the DTV, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive to RFI covering such a broad spectrum.
